---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# track2dm

<!-- badges: start -->
<!-- badges: end -->

The goal of track2dm is to create detection matrix from transect lines that account for topography variability. The detection matrix is usually used for hierarchical modelling of species occupancy and distribution (see MacKenzie et al., 2004 about occupancy modelling).

## Installation

You can install the released version of track2dm and the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ilubis85/track2dm")
```
## Example

We have provided some dataset to play with to understand how the package works. Three types of data are needed, the track where the survey is recorded spatially, the observation along the tracks and elevation as raster data downloaded from USGS website.

```{r example}
# LOAD ALL DATA
# Read elevation raster data from the package
data("elevation")

# Read a survey track from the package
data("track")

# Read the observation from the package
data("observation")

# Convert dataframe (track and observation to spatial points data-frame
track_pt <- track2dm::df2sp(track, UTMZone = "+proj=utm +zone=47 +datum=WGS84 +units=m +no_defs")
observation_pt <- track2dm::df2sp(observation, UTMZone = "+proj=utm +zone=47 +datum=WGS84 +units=m +no_defs")

```

Track is a dataframe contains date and time, X, Y and usually Z information downloaded from GPS. Whilst observation is a dataframe contains any information about the observed species. The elevation is needed to extract Z values to calculate distance in three dimension (3D).

```{r track}
head(track, 5)
head(observation, 5)
```

This is the data look like when we plot them using tmap package:

```{r elevation, echo = FALSE}
library(tmap)
tm_shape(elevation) +
  tm_raster(style = "cont") +
  tm_shape(track_pt) + tm_dots(col = "grey", shape=20, size=0.5) +
  tm_shape(observation_pt) + tm_dots(col = "red", shape=16, size=0.6)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
